Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
small changes

{\tt edgeR} is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero. Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.

@@

Gene expression heterosis is speculated to be one possible explanation for phenotypic
heterosis of traits like plant height or grain yield. Existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis due to the composite null hypothesis in equation \eqref{e:hypotheses}. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses. We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be done efficienctly via parallelization. This methodology can be conveniently implemented in the statistical software {\tt Stan} which also allows for much more complex models. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.

>>>>

The {\tt edgeR} approach is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero. Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.

@@

Gene expression heterosis is speculated to be one possible explanation for phenotypic
heterosis of traits like plant height or grain yield. Existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis due to the composite null hypothesis in equation \eqref{e:hypotheses}. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses. We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be accomplished efficiently via parallelization. This methodology can be conveniently implemented in the statistical software {\tt Stan} which also allows for much more complex models. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
